\texttt{MSnbase}, efficient R-based access and manipulation of raw mass spectrometry data

### 2020-04-25 msnbase2.Rnw
compose one bench figure

The former is the legacy storage mode, implemented in the first
version of the package, that loads all the raw data and the metadata
in memory. This solution doesn't scale for modern large dataset, and
was complemented by the on-disk backend, that only load metadata in
memory and accesses the spectra in the original files when
needed. There are two direct benefits using the on-disk
backend. Figure \ref{fig:bench1} shows 4-fold faster reading times
(left) and 10-fold smaller memory footprint (right).

@@

p1 <- data.frame(time) %>% 
    rownames_to_column() %>% 
    pivot_longer(names_to = "files", 
                 values_to = "time", 
                 -rowname) %>% 
    rename(backend = rowname) %>% 
    mutate(files = as.numeric(sub("X", "", files))) %>% 
    ggplot(aes(x = files, y = time, colour = backend)) + 
    geom_point() + 
    geom_line() + 
    scale_x_continuous(breaks=c(1, 5, 10)) + 
    ylab("Reading time (s)") + 
    theme(legend.position = "none")

p2 <- data.frame(sz) %>% 
    rownames_to_column() %>% 
    pivot_longer(names_to = "files", 
                 values_to = "size",
                 -rowname) %>% 
    rename(backend = rowname) %>% 
    mutate(files = as.numeric(sub("X", "", files))) %>% 
    ggplot(aes(x = files, y = size, colour = backend)) + 
    geom_point() + 
    geom_line() + 
    scale_x_continuous(breaks=c(1, 5, 10)) + 
    ylab("Object size (MB)") 
@

\begin{figure}[ht]
  \centering
<<plot_bench1, out.width = '100%', fig.width = 7, fig.height = 3>>=
p1 + p2 + plot_annotation(tag_levels = 'a')
@
\caption{(a) Reading time (in seconds) and (b) data size in memory (in
  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
  only) and 6103 MS2 (on-disk and in-memory) spectra. }
\label{fig:bench1}
\end{figure}

@@

As shown on figure~\ref{fig:filt}, this lazy mechanism si
significantly faster than its application on in-memory data.

\begin{figure}[ht]
  \centering
<<plot_filter, message = FALSE, out.width = '50%', fig.width = 5, fig.height = 3>>=
load("bench_t_filt.rda")
ggplot2::autoplot(t_filt)
@
\caption{Filtering benchmark assessed over 10 interations on in-memory (top) and on-disk (bottom) data containing 6103 MS2 spectra.}
\label{fig:filt}
\end{figure}

The advantageous reading and execution times and memory footprint of
the on-disk backend are possible by avoiding unnecessary access to the
raw data. Once access to the spectra M/Z and intensity values become
mandatory (for example for plotting), then the in-memory backend
becomes more efficient, as illustrated on
figure~\ref{fig:access}. This gain is maximal when the whole dataset
is the be accessed (i.e. all spectra are already in memory) and
negligeable when large fractions of the data need to be subset.

\begin{figure}[ht]
  \centering
<<plot_access, message = FALSE, out.width = '75%', fig.width = 5, fig.height = 3>>=
load("bench_t_access.rda")
tibble(time = microbenchmark:::convert_to_unit(t_access$time, "ms"),
       expr = as.character(t_access$expr)) %>%
    mutate(n = sub("^access_.+_", "", expr)) %>%
    mutate(n = sub("all", 6103, n)) %>%
    mutate(mode = if_else(grepl("mem", expr), "in-memory", "on-disk")) %>%
    ggplot(aes(x = n, y = time, fill = log10(as.numeric(n)))) + 
    ggplot2::geom_violin() + 
    ggplot2::scale_y_log10() +
    ylab("Time [millisec]") +
    xlab("Number of spectra (out of 6103)") +
    facet_wrap(~ mode) +
    theme(legend.position = "none")
@
\caption{Access time to spectra for the in-memory (left) and on-disk
  (right) backends for 1, 10, 100 1000, 5000 and all 6103 spectra.}
\label{fig:access}
\end{figure}

>>>>

@@

p_read <- 
    data.frame(time) %>% 
    rownames_to_column() %>% 
    pivot_longer(names_to = "files", 
                 values_to = "time", 
                 -rowname) %>% 
    rename(backend = rowname) %>% 
    mutate(files = as.numeric(sub("X", "", files))) %>% 
    ggplot(aes(x = files, y = time, colour = backend)) + 
    geom_point() + 
    geom_line() + 
    scale_x_continuous(breaks=c(1, 5, 10)) + 
    ylab("Reading time (s)") + 
    xlab("Number of files") +
    theme(legend.position = "none")

p_sz <- 
    data.frame(sz) %>% 
    rownames_to_column() %>% 
    pivot_longer(names_to = "files", 
                 values_to = "size",
                 -rowname) %>% 
    rename(backend = rowname) %>% 
    mutate(files = as.numeric(sub("X", "", files))) %>% 
    ggplot(aes(x = files, y = size, colour = backend)) + 
    geom_point() + 
    geom_line() + 
    scale_x_continuous(breaks=c(1, 5, 10)) + 
    ylab("Object size (MB)") +
    xlab("Number of files") +
    theme(legend.position = "none")    
@

<<filt>>=
load("bench_t_filt.rda")
p_filt <- 
    tibble(time = microbenchmark:::convert_to_unit(t_filt$time, "ms"),
           expr = as.character(t_filt$expr)) %>%    
    mutate(mode = if_else(grepl("mem", expr), "in-memory", "on-disk")) %>%
    ggplot(aes(x = mode, y = time, fill = mode)) + 
    ggplot2::geom_violin() + 
    ggplot2::scale_y_log10() +
    ylab("Filtering time [millisec]") + 
    xlab("Backend") +    
    theme(legend.position = "none")

@

<<access>>=
load("bench_t_access.rda")
p_access <- 
    tibble(time = microbenchmark:::convert_to_unit(t_access$time, "ms"),
           expr = as.character(t_access$expr)) %>%
    mutate(n = sub("^access_.+_", "", expr)) %>%
    mutate(n = sub("all", 6103, n)) %>%
    mutate(mode = if_else(grepl("mem", expr), "in-memory", "on-disk")) %>%
    ggplot(aes(x = n, y = time, fill = mode)) + 
    ggplot2::geom_violin() + 
    ggplot2::scale_y_log10() +
    ylab("Raw data access [millisec]") +
    xlab("Number of spectra (out of 6103)") +
    facet_wrap(~ mode) +
    theme(legend.position = "none")
@

\begin{figure}[p]
  \centering
<<plot_bench>>=
(p_read + p_sz + p_filt) / p_access + 
    plot_layout(heights = c(1, 0.75)) + 
    plot_annotation(tag_levels = 'a') &
    theme(axis.title = element_text(size = 10))
@
\caption{(a) Reading time (in seconds) and (b) data size in memory (in
  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
  only) and 6103 MS2 (on-disk and in-memory) spectra. (c) Filtering
  benchmark assessed over 10 interations on in-memory and on-disk data
  containing 6103 MS2 spectra.  (d) Access time to spectra for the
  in-memory (left) and on-disk (right) backends for 1, 10, 100 1000,
  5000 and all 6103 spectra. On-disk backend: blue. In-memory backend:
  red. }
\label{fig:bench}
\end{figure}

The former is the legacy storage mode, implemented in the first
version of the package, that loads all the raw data and the metadata
in memory. This solution doesn't scale for modern large dataset, and
was complemented by the on-disk backend, that only load metadata in
memory and accesses the spectra in the original files when
needed. There are two direct benefits using the on-disk
backend. Figure \ref{fig:bench} shows 4-fold faster reading times (a)
and 10-fold smaller memory footprint (b).

@@

As shown on figure~\ref{fig:bench} (c), this lazy mechanism si
significantly faster than its application on in-memory data. The
advantageous reading and execution times and memory footprint of the
on-disk backend are possible by avoiding unnecessary access to the raw
data. Once access to the spectra M/Z and intensity values become
mandatory (for example for plotting), then the in-memory backend
becomes more efficient, as illustrated on figure~\ref{fig:bench}
(d). This gain is maximal when the whole dataset is the be accessed
(i.e. all spectra are already in memory) and negligeable when large
fractions of the data need to be subset.

### 2020-04-25 msnbase2.tex
compose one bench figure

The former is the legacy storage mode, implemented in the first
version of the package, that loads all the raw data and the metadata
in memory. This solution doesn't scale for modern large dataset, and
was complemented by the on-disk backend, that only load metadata in
memory and accesses the spectra in the original files when
needed. There are two direct benefits using the on-disk
backend. Figure \ref{fig:bench1} shows 4-fold faster reading times
(left) and 10-fold smaller memory footprint (right).

\begin{figure}[ht]
  \centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=1\linewidth]{figure/plot_bench1-1}

\end{knitrout}
\caption{(a) Reading time (in seconds) and (b) data size in memory (in
  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
  only) and 6103 MS2 (on-disk and in-memory) spectra. }
\label{fig:bench1}
\end{figure}

@@

As shown on figure~\ref{fig:filt}, this lazy mechanism si
significantly faster than its application on in-memory data.

\begin{figure}[ht]
  \centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=0.5\linewidth]{figure/plot_filter-1}

\end{knitrout}
\caption{Filtering benchmark assessed over 10 interations on in-memory (top) and on-disk (bottom) data containing 6103 MS2 spectra.}
\label{fig:filt}
\end{figure}

The advantageous reading and execution times and memory footprint of
the on-disk backend are possible by avoiding unnecessary access to the
raw data. Once access to the spectra M/Z and intensity values become
mandatory (for example for plotting), then the in-memory backend
becomes more efficient, as illustrated on
figure~\ref{fig:access}. This gain is maximal when the whole dataset
is the be accessed (i.e. all spectra are already in memory) and
negligeable when large fractions of the data need to be subset.

\begin{figure}[ht]
  \centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=0.75\linewidth]{figure/plot_access-1}

\end{knitrout}
\caption{Access time to spectra for the in-memory (left) and on-disk
  (right) backends for 1, 10, 100 1000, 5000 and all 6103 spectra.}
\label{fig:access}
\end{figure}

>>>>

\begin{figure}[p]
  \centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/plot_bench-1}

\end{knitrout}
\caption{(a) Reading time (in seconds) and (b) data size in memory (in
  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
  only) and 6103 MS2 (on-disk and in-memory) spectra. (c) Filtering
  benchmark assessed over 10 interations on in-memory and on-disk data
  containing 6103 MS2 spectra.  (d) Access time to spectra for the
  in-memory (left) and on-disk (right) backends for 1, 10, 100 1000,
  5000 and all 6103 spectra. On-disk backend: blue. In-memory backend:
  red. }
\label{fig:bench}
\end{figure}

The former is the legacy storage mode, implemented in the first
version of the package, that loads all the raw data and the metadata
in memory. This solution doesn't scale for modern large dataset, and
was complemented by the on-disk backend, that only load metadata in
memory and accesses the spectra in the original files when
needed. There are two direct benefits using the on-disk
backend. Figure \ref{fig:bench} shows 4-fold faster reading times (a)
and 10-fold smaller memory footprint (b).

@@

As shown on figure~\ref{fig:bench} (c), this lazy mechanism si
significantly faster than its application on in-memory data. The
advantageous reading and execution times and memory footprint of the
on-disk backend are possible by avoiding unnecessary access to the raw
data. Once access to the spectra M/Z and intensity values become
mandatory (for example for plotting), then the in-memory backend
becomes more efficient, as illustrated on figure~\ref{fig:bench}
(d). This gain is maximal when the whole dataset is the be accessed
(i.e. all spectra are already in memory) and negligeable when large
fractions of the data need to be subset.
